# Preparing the Data {#Prepping}

In this section, we are using R to prepare data for analysis. We will use mock data from the TIMSS Teacher Questionnaire.

## Import the data from the CSV file

I generated these mock data as fake responses to the TIMSS math teacher questionnaire https://nces.ed.gov/timss/pdf/2015_8th_grade_Teacher_Questionnaire_Math.pdf. You can also get real data from their site. We could use the code below with those data (though we would need to address missing values). In other words, this can be part of our data analysis plan and we created data using our best guess about how people would respond. Then, we assume it functions well, and then analyze the real data.

If you've opened this RStudio file from the same folder that contains the CSV file, `Week13_Timss_Mock_Raw.csv`, your working directory will be that location and you can import data using this code:

```{r}
timss <- read.csv('Week13_Timss_Mock_Raw.csv') 
```

If you prefer to download the data, it will be temporarily available here:

```{r, eval=FALSE}
timss <- read.csv("http://www2.hawaii.edu/~georgeha/Handouts/Week13_Timss_Mock_Raw.csv")
```

### Examining the imported data set

We can print the first and last five rows of data using the `head()` and `tail()` functions:
```{r eval = FALSE}
head(timss)
tail(timss)
```

We can also select which rows to print using indexing. Here, it prints only rows 1, 3, 5, 7, and 9: 

```{r, eval=FALSE}
timss[c(1,3,5,7,9), ]
```

We can also print the data set to console or view it in a new tab. 

```{r eval = FALSE}
timss
View(timss)
```

Let's check that the data set has the number of observations we expect. We should have 30 mock teachers. This prints the number of rows in the column labeled `TID`.

```{r}
length(timss$TID)
```

Good. There are 30. We can also see what how many rows and columns there are with the `dim()` function, which prints the number of rows then the number of columns.

```{r}
dim(timss)
```

Let's see what the names of the variables are:

```{r}
names(timss)
```

If we enclose this code in the `dput()` function, we can then copy and paste names into code.
```{r}
dput(names(timss)) 
```

This will be handy if we wish to copy and paste a string of column names from our console into our code because it formats it with quotes around the names and commas in between them, which is what we want to use in the `c()` function.

## Managing the structure of the data
Let's look at the structure of our data set, using the `str()` function. 

```{r}
str(timss)
```

We can see that Age and Degree are numeric. We see that the Q6 demographic questions are "character", which is appropriate for data that are text in CSV file. The other variables are numeric (integer).^[In earlier versions of R, the `read.csv()` function read these as factors, which are appropriate for nominal and ordinal level data.]

Let's change a character variable to numeric data. For the Q6 variables, there are "No" and "Yes" responses. These are **not** numbers, so we cannot perform mathematical functions on them. We can create a new dichotomous variable from these one or more nominal variables, just like how we did in Excel. Let's create a `Major_in_Educ` variable using the `ifelse()` function. The vertical bar, |, symbolizes "or", so the following code is _if EdMath is `Yes`, or if EdScience is `Yes`, or if EdGen is `Yes`, code the new variable as `1`; otherwise, code it as `0`_.

```{r}
timss$Q6_Major_in_Edu <- ifelse(timss$Q6_EdMat == "Yes" |
                                timss$Q6_EdSci == "Yes" |    
                                timss$Q6_EdGen == "Yes", 1 ,0)
```

Let's verify it worked by examining some rows of data---here, the first 12 rows and the columns that were used to get our new variable as well as the new variable:

```{r}
timss[1:12, c("TID", "Q6_EdMat", "Q6_EdSci", "Q6_EdGen", "Q6_Major_in_Edu")]
```

The variable we just created is in the rightmost column. We can look at the structure of this selected set of columns and see that the last one is numeric. 
```{r}
str(timss[, c("Q6_EdMat", "Q6_EdSci", "Q6_EdGen", "Q6_Major_in_Edu")])
```
This dummy coding is valuable for analysis. We have prepared this new variable so it can serve as a predictor in a regression model (i.e., a general linear model).

## Recoding response-category data
### Reverse-coding and rescaling data

Questions 15a--15g have numeric codes that are not logical. For instance, _Every or almost every lesson_ is $1$ but _Never_ is coded as $4$. Usually, we use higher values for more of a construct and lower values for less. 

So, let's change our coding so that _Never_ is coded as $0$, _Some lessons_ is $1$, _About half the lessons_ is $2$, and _Every or almost every lesson_ is $3$. So, we are going from a 4-to-1 scale to a 0-to-3 scale.

The general formula for reverse coding is this:

 * newvalue = (num of response options) + (bottom value of new scale) - (observed value in the original scale)

So with our data in Questions 15a--15g, 

 * How many response categories are there? _Answer:_ 4
 * What's the bottom value on our NEW scale going to be? _Answer:_ 0
 * What are our original data? _Answer:_ all of the data in Q15a through Q15g

Maybe we forgot the names of the variables we were investigating. Let's use `names()` to see what's available.

```{r}
names(timss) 
```

Let's try the calculation on a single variable and add "R" to the new variable name so we can remember this is the reversed version:

```{r}
timss$Q15_StCentrd_aR <- 4 + 0 - timss$Q15_StCentrd_a
```

Let's verify it worked by comparing the reversed and the original data with first 10 rows.
Any old 4s should now be $0$ (because 4 + 0 - 4 = 0), 3s should be $1$, 2s should still be $2$, and 1s should be $3$.

```{r}
timss[1:10, c("Q15_StCentrd_a","Q15_StCentrd_aR")]
```

Let's do this same operation with the remaining 6 items in Q15 (Q15b through Q15g). We could repeat the line above with the new variable names in each line. Or, we can create an R object that contains the names of these 7 variables and then perform the same calculation on this object:

**Step 1.** Let's create an object that contains the 7 items in Question 15. We'll call this Q15.^[We can shorten this code using the `paste0()` function and the first 7 letters of the alphabet: `Q15 <- paste0("Q15_StCentrd_", letters[1:7])`]

```{r}
Q15 <- c("Q15_StCentrd_a", "Q15_StCentrd_b", "Q15_StCentrd_c", "Q15_StCentrd_d", 
        "Q15_StCentrd_e", "Q15_StCentrd_f", "Q15_StCentrd_g")
```

This object, Q15, is only a list of names, but we can use it to refer to those data.^[The benefit here is that we can re-use this object, Q15, in other code to reference these columns.]

```{r}
Q15
```

Let's use our object, now: Here are the first five people's responses on these items:

```{r}
timss[1:5, Q15]
```

Let's check the range of the original data. The minimum should be 1 and the maxium 4: 

```{r}
range(timss[ , Q15])
```

So, we have used the object `Q15` in two different functions. This is better than retyping this long string: `c("Q15_StCentrd_a", "Q15_StCentrd_b", "Q15_StCentrd_c", "Q15_StCentrd_d", 
        "Q15_StCentrd_e", "Q15_StCentrd_f", "Q15_StCentrd_g")`.

#### Reverse coding

**Step 2.** Let's do the same thing with the new set of variables that we will create. Let's use an "R" to stand for Reversed. We could write out all the names, with R after each variable, or we can take a shortcut and use the `paste0()` function to concatenate an "R" at the end of every element in the Q15 object:

```{r}
Q15R <- paste0(Q15,"R")
```
Verify that it created the set of names we want:
```{r}
Q15R
```

Let's create these seven new reverse-coded variables using our formula and these two objects, Q15 and Q15R. Because each object includes the same number of names, it performs the calculation on the first variable, then the next one and so forth.

```{r}
timss[ ,Q15R] <- (4 + 0 - timss[,Q15]) 
```
Take a look at the first 6 rows:

```{r}
timss[1:6, Q15]
timss[1:6, Q15R]
```
These match our Excel calculations.

Let's look at the first variable side by side:

```{r}
timss[1:6, c(Q15[1], Q15R[1]) ] 
```

We used the index [1] to ask for the first element of the Q15 and Q15R objects. This is the same as if we asked for the names:

```{r}
timss[1:6, c("Q15_StCentrd_a", "Q15_StCentrd_aR") ]
```

We could ask for the second element after making a small change to the code:

```{r eval = FALSE}
timss[1:6, c(Q15[2], Q15R[2]) ] 
```

Let's see whether the minimum and maximum values are outside the bounds of 0, and max is 3.

```{r}
range(timss[,Q15R]) 
```

We could also look at each of the 7 questions:

```{r}
apply(timss[ , Q15R], 2, FUN = range)
```

None of the values is lower than 0 or higher than 3, which is what we expected.

#### Rescaling Q16 so zero is meaningful
Question 16 has the numeric codes in a logical direction (where a smaller numeric value represents a lower amount on the concept), but let's change the scale from a 1-to-3 scale to 0-to-2 so that "Not at all" is zero.^[This recoding is helpful if we use these as predictors in a regression because we can interpret the intercept as a meaningful number.] For this, we have to subtract 1 from the original value.

Let's remember what these Q16 variable names are:^[The `dput()` function creates an output that is easy to copy and paste into the `c()` function in our next step.]
```{r}
dput(names(timss)) 
```

Let's assign them to the object Q16:
```{r}
Q16 <- c("Q16_Lims_a", "Q16_Lims_b", "Q16_Lims_c", "Q16_Lims_d", "Q16_Lims_e", 
        "Q16_Lims_f", "Q16_Lims_g")
range(timss[ , Q16])
```

Because Q16 only needs to be lowered so that _Not at all_ = $0$, we might use "Z" in the names to remind us they are rescaled to have a meaningful zero.
```{r}
Q16Z <- paste0(Q16,"Z")
Q16Z 
```

Now, let's subtract 1 from the original response and create our new variables in the data frame:
```{r}
timss[ ,Q16Z] <- (timss[ ,Q16] - 1) 
range(timss[,Q16Z])
```

We can see these new variables are appended at the end of the data set:
```{r }
str(timss) 
```

Let's check that it worked with the first and 7th variable.^[We could also copy and paste the code and replace the numbers with any within the range of 1 through 7 because there are 7 items in this question.]

```{r eval=FALSE}
timss[7:12, c(Q16[1], Q16Z[1])]
timss[7:12, c(Q16[7], Q16Z[7])]
```

#### Rescaling Q19
Similar to Question 15, Question 19's response scale categories are in an illogical direction. Let's do the same thing we did for Question 15.

```{r eval = FALSE}
str(timss) 

Q19 <- c("Q19_Didac_a",  "Q19_Didac_b",  "Q19_Didac_c")
Q19R<- c("Q19_Didac_aR", "Q19_Didac_bR", "Q19_Didac_cR") 

timss[,Q19R] <- (4 + 0 - timss[,Q19]) 
```


```{r echo = FALSE}
Q19 <- c("Q19_Didac_a",  "Q19_Didac_b",  "Q19_Didac_c")
Q19R<- c("Q19_Didac_aR", "Q19_Didac_bR", "Q19_Didac_cR") 

timss[,Q19R] <- (4 + 0 - timss[,Q19]) 
```


## Creating Composite Scores (AKA Indexes)

**1)** Let's create an index (AKA scale or composite) variable from the 7 numeric items in Q15. Let's name our new variable `Q15_Stdnt_Centred_Index`.

```{r}
timss$Q15_Index <- apply(timss[ ,Q15R],1, mean, na.rm = TRUE) 
```

Applying the mean function across columns "1" directs R to apply this function to the first margin, i.e. to the rows (and across columns). See http://www2.hawaii.edu/~georgeha/surveys/_book/appending.html.
So we get each person's mean score across the 7 items in Q15R. `na.rm = TRUE` is used in case we have any missing data.

Let's look at the index scores for the first 5 or so observations:
```{r}
timss[1:5, c(Q15R,"Q15_Index")]  
summary(timss$Q15_Index)
```

We can use the structure function, `str()`, with the same subset of data:
```{r}
str(timss[, c(Q15R,"Q15_Index")])
```

**2)** Let's do the same thing with Q16 and Q19:
```{r}
timss$Q19_Index <- apply(timss[,Q19R], 1, mean, na.rm = TRUE)
timss$Q16_Index <- apply(timss[,Q16Z], 1, mean, na.rm = TRUE)
```


Take a look at the new composite variables:
```{r}
timss[1:5,c(Q19R,"Q19_Index")]  
timss[1:5,c(Q16Z,"Q16_Index")]  
```



## A general formula for rescaling data

Here's a more general formula for rescaling:

Sometimes, we want to rescale data. For example, if we want to create an index based on several variables, but some of these variables are on a 1-to-5 scale while others are on a 1-to-6 scale and we want to give equal weight to the variables, we would rescale one type to be on the other's scale.

We can use this general formula, with the "bottom" and the "top" referring to the bottom and top of the scale and the "value" referring to the response of interest (a person's response): new.value = new.bottom + (old.value - old.bottom)*(new.top - new.bottom)/(old.top - old.bottom) but first we have to assign values to each of the objects on the right side (or we could just plug them in). Here is an example of rescaling a response of 3 that was on a scale of 1 to 6 to the new scale of 1 to 5:

```{r}
old.value <-  3  #This is our original data point. 
old.bottom<-  1  #This is the lowest possible point on the original scale
old.top   <-  6  #This is the highest possible point on the original scale
new.bottom<-  1  #This is the lowest possible point on the new scale
new.top   <-  5  #This is the highest possible point on the new scale
```

```{r}
new.bottom + (old.value - old.bottom)*(new.top - new.bottom)/(old.top - old.bottom)
```
which is the same as this if we weren't using object names:
```{r}
1 + (3 - 1)*(5-1)/(6-1)
```
Seems like too much work, right. However, we can use this in a custom function.

## Create and use a custom function

It's convenient to create a custom function using that formula. Let's name our custom function something that does not already exist in base R. We can check this using the help() function to see if something comes up:

```{r eval = FALSE}
help("myrescale") 
```
Seems like "myrescale" is a good name because nothing comes up in help. After the function(), we specify the argument names and their order. Then, we enclose the operations in the curly brackets. The final object, `new.value`, is what is produced.

```{r}
myrescale <- function(old.value, old.bottom, old.top, new.bottom, new.top){
 new.value <- new.bottom + (old.value - old.bottom)*(new.top - new.bottom)/(old.top - old.bottom)
}
```

Now we can use that custom function. Let's make up some arbitrary original data,
within our original 1 to 6 scale: 

```{r}
olddat <- c(4, 3, 2, 1, 6, 5)

new1to5 <- myrescale(old.value  = olddat, 
                     old.bottom = 1, 
                     old.top    = 6, 
                     new.bottom = 1, 
                     new.top    = 5)
new1to5
```

So, we used our homemade function, `myrescale()`, by specifying the arguments. We can re-use this function in our R session.

Here's another example with our data set: 
Maybe for some reason we want to transform all of our Q15R variables to a 5 point scale that is centered at the midpoint, so from -2 to 2:

First let's create a name for our new set of variables Maybe we use "C5" to remind ourselves it's centered and on five points:

```{r}
Q15Centr5 <- paste0(Q15, "_C5")
Q15Centr5

timss[ ,Q15Centr5]<- myrescale(old.value  = timss[, Q15R], 
                      old.bottom =  0, 
                      old.top    =  3, 
                      new.bottom = -2, 
                      new.top    =  2)
timss[1:6, c(Q15R[3], Q15Centr5[3]) ]
```

We can use this general formula to reverse scale, too. To reverse the scale, the new top and bottom are just reversed:

```{r}
newReversed <- myrescale(olddat, 
                         old.bottom = 1, 
                         old.top    = 6, 
                         new.bottom = 6, 
                         new.top    = 1)
```

Let's compare our newly reversed scale to our original scale:
```{r}
newReversed
olddat
```

## The code in one place for one of the composite scores
Here is the code again for one of the scales---the Q15 index---but without the extra steps above. This shows the core steps:

 1. Reading in the CSV file,
 2. Examining the structure of the data frame,
 3. Assigning the names of the columns to an object, `Q15`, so we can use that object in subsequent code,
 4. Creating a new object, `Q15R` that includes the names of the `Q15` items but with an "R" so we can create a new set of columns that are rescaled,
 5. The actual recscaling of the items,
 6. Calculating the composite score from the rescaled items, and
 7. Examining the structure of our rescaled items and our composite score.
```{r results = "hide", eval = FALSE}
timss <- read.csv('Week13_Timss_Mock_Raw.csv') 
str(timss)
Q15   <- c("Q15_StCentrd_a", "Q15_StCentrd_b", "Q15_StCentrd_c", "Q15_StCentrd_d", 
        "Q15_StCentrd_e", "Q15_StCentrd_f", "Q15_StCentrd_g")
Q15R  <- paste0(Q15,"R")
timss[ ,Q15R]   <- (4 + 0 - timss[,Q15]) 
timss$Q15_Index <- apply(timss[ ,Q15R],1, mean, na.rm = TRUE) 
str(timss[, c(Q15R,"Q15_Index")])
```

## A shortcut, when all of the items are on the same scale

So far, we have been practicing rescaling items before we use them to create a composite score. That was worth learning because sometimes we find ourselves needing to recode only some of the items (e.g., if we need to reverse-code items that are phrased in the opposite direction to the other items in the scale). 

When we do not have that situation, we can take a simpler approach: We can create the composite variable and then rescale that. Here's an example with Q15.

```{r}
Q15 <- paste0("Q15_StCentrd_", letters[1:7])
timss$Q15_Index_raw  <- apply(timss[ ,Q15],1, mean, na.rm = TRUE) 
timss$Q15_Index_easy <- 4 + 0 - timss$Q15_Index_raw
```

Annoying, huh. We did all of the work above when we could have achieved the same result with three lines of code. Just to verify, in this demo, let's eyeball the first 10 people's scores on this variable (timss$Q15_Index_easy) and the one we had created earlier using after rescaling the individual items (Q15_Index):

```{r}
timss[1:10, c("TID", "Q15_Index", "Q15_Index_easy")]
```

## Summary
In this chapter, we created variables (dummy codes and composite scores), recoded and rescaled variables (items and composite variables), created a custom function, and practiced looking at our data and the structure of our data.



