# Installing and Using Packages {#Packages}

In R, we use packages that are developed by nerdy people (whom I admire) and made available for the rest of us to use. Often a package is for a specific purpose or a family of analyses. To analyze simple survey data, we will use the psych package [@R-psych] to perform descriptive statistics and estimate reliability coefficients, the ggplot2 package [@R-ggplot2] to create pretty plots, and the dplyr [@R-dplyr] package to wrangle our data.

## Seeing if we need to install a package

First, we need to see if the packages we need for this session are installed on our machine. To do this, we can either

 1. see if each of these packages is listed in the "Packages" tab in the lower right pane, or
 2. use the `require()` function with each package to see if we get the result "there is no package called '[package name]' ". Let's say we think there's a package called PackageNot on our machine. (You can plug in any random name for this.) We can verify this with this code:

```{r}
require("PackageNot")
```

Do the same thing with the psych package. If you do _not_ have it installed, it will give you the same result. If you do have it installed, it will generate an output that says `Loading required package: psych`:
```{r}
require("psych")
```

You might also see a warning that the package was built under a newer version of R. There are R updates every few months or so. If you find the package is not working, you might need to update R, which we won't address right now.

## Installing a package

The `install.packages()` function is handy for installing packages:

```{r eval = FALSE}
install.packages("psych")
```

Once a package is installed on a machine, we do not need to install it again.^[It is a good idea, however, to update the packages every once in a while. In RStudio, we can click on the Packages tab and click Update.] But, you might want to keep your code as a record of what you did or make the `install.packages()` line of your code accessible to your friends if you wish to share it with others.^[Talking about R functions in social situations is a great way to ensure you get more alone time.] You can comment out the code if you want R to ignore it. In R, anything after a `#` symbol is commented out:

```{r eval = FALSE}
# install.packages("psych")
```

Now, let's bring the package into our working library so we can use a function from that package.^[Base R has a lot of inconsistencies. You may notice that we need quotes around the package name when we use the `install.packages()` function but we don't when we use the `library()` function. This is part of the so-called fun in R.]
```{R message=FALSE}
library(psych)
```

We can also find out what functions are available in a package with the `help()` function and the package name:
```{r eval = FALSE}
help(package = "psych")
```
The functions are in alphabetical order. We can see the `alpha()` function listed second in the alphabet (under `-- A --`). We will use this function for estimating internal consistency reliability. If you click on the function, there will be  a lot of information to sort through. The `Usage`and `Arguments` sections near the top and the `Examples` section at the bottom are usually useful.

## Using a function from a package

After we bring the package into our library, during our R session, we can use its functions. 

One that we will use is the `describe()` function. First, let's review some code that we learned last week:^[Also, note that because all of our items require the same rescaling procedure, we could use a shorter version of this code, discussed at the end of the last chapter.]

```{r }
timss <- read.csv('Week13_Timss_Mock_Raw.csv') 

Q15 <- c("Q15_StCentrd_a", "Q15_StCentrd_b", "Q15_StCentrd_c", "Q15_StCentrd_d", 
        "Q15_StCentrd_e", "Q15_StCentrd_f", "Q15_StCentrd_g")

Q15R <- paste0(Q15,"R")

timss[ ,Q15R] <- (4 + 0 - timss[,Q15]) 

timss$Q15_Index <- apply(timss[ ,Q15R],1, mean, na.rm = TRUE) 

```

We could use the `summary()` function from Base R to get some descriptive statistics of this variable, Q15_Index.

```{r}
summary(timss$Q15_Index)
```

This does not show the standard deviation or the _n_-size. Now, let's try the `describe()` function from the psych package:

```{r}
describe(timss$Q15_Index)
```

With this, we get the _n_-size (`n`), the standard deviation (`sd`), and some other handy estimates, such as the standard error of the mean (`se` in this output). We might not wish to report all of these numbers from the output, but this is certainly more informative than Base R's `summary()` function.







