---
title: "Getting Started with R, Part 2"
subtitle: "EDEP 602, Survey Research Design & Analysis"
author: "George M. Harrison"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

##Learning Objectives
1. Import data from a CSV file.  
1. Examine the data
1. Summarize the data

##Importing Data from a CSV file
Our data set contains 10 observations. They are school professionals who responded to a questionnaire that had five Likert-type questions and one yes-no question asking whether or not they are a teacher.
Here are the data: 
```{marginfigure}
_ID_ is the respondent's identifier code, which we arbitrarily assigned. _Lik1_ through _Lik5_ are our Likert-type items, corresponding to the five questions. _Teacher_ is coded _Yes_ for _yes, this person is a teacher_ and _No_ for _no, they're not_.
```
```{r, include=FALSE}
dat<- read.csv("SchoolProfessionalSurvey.csv")
```
```{r, echo=FALSE}
knitr::kable(dat,align=rep('c',7))
```

Let's read the data set into R and save it to an object, which we'll arbitrarily call _dat_. The `read.csv()` function does just that---it reads CSV files. Be sure that the CSV file is in the same folder as your .r file (unless you're a few steps ahead and have set the working directory). Also, be sure to use quotes around the name. R is __case sensitive__, so if you get an error, check the spelling. Let's also look at the first several lines of data by performing the `head()` function on the object we just created:`r margin_note("The less-than symbol '<' and the dash symbol '-' together as '<-' are the 'assign to' operation in R. We have instructed R to read the CSV file that is in our currect directory (or folder) and assign that to an object labeled 'dat' in our R session.")`
`r margin_note("When you run this code, you'll see the output in your console. In this handout, it is displayed with double hash marks preceding each line.")`
```{r}
dat<- read.csv("SchoolProfessionalSurvey.csv")
head(dat)
```
We can also examine the last six observations of our data using the `tail()` function.`r margin_note("I suppose you can say we're making heads and tails of our data.")`
```{r}
tail(dat)
```

##Describing Our Data
The `summary()` function is convenient. We can describe all of the variables in a data set^[A data set is called a ___data frame___ in R.] using the function on our data frame object, dat:

`r margin_note("Notice that we have also summarized the ID variable, which is not meaningful. We would obviously not include that in a report.")`
```{marginfigure}
Notice that _Teacher_ is not a numeric variable. Notice also that the results of the summary function do not return the mean, median and so forth because those would not make sense with non-numeric data. Instead, the frequency of observations in each category of this variable is reported.
```
```{r}
summary(dat)
```
We can apply the `summary()` function to a single variable:
`r margin_note("The dollar sign is used to select a particular column (i.e., variable) from the data. All rows of this column of data are included.")`
```{r}
summary(dat$Lik1)
```
We can also perform functions like `min()`, `max()`, `mean()` and assign them to objects that we can later use:`r margin_note("For each of the three lines, we performed the function on the variable and we assigned the output to an object, which we've arbitrarily named  minX, maxX, and meanX.")`
```{r}
minX <- min(dat$Lik1)
maxX <- max(dat$Lik1)
meanX<- mean(dat$Lik1)
```
Let's view the objects in the console:
```{r}
meanX
minX
maxX
```
There are many functions we can use. For example, we can get the variance of a variable using the `var()` function and the standard deviation using `sd()`:
```{r}
var(dat$Lik1)
sd(dat$Lik1)
```










