
---
title: "Introduction to R"
subtitle: "EDEP 602"
author: "George M. Harrison"
date: "`r Sys.Date()`"
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```






###Learning Objectives
* Describe variables  
* Calculate the covariance and correlation between two variables  
* Calculate the covariance matrix and correlation matrix between many variables  
* Create a bivariate scatterplot (correlation)  
* Fit a simple linear regression model  
* Predict Y from X using our model  

###Data
Our data set contains 10 observations, ID 1:10. They are students who took a five-item test and who were rated by their teacher on five five-point Likert-type items.  We might be interested in whether the teacher was able to predict the students' test scores.
Let's read the data in and save it to an object, which we'll call dat. Let's look at the data, too:
```{r}
dat<- read.csv("Practice2Data.csv")
dat
summary(dat)
```
Describing Our Data

The `summary()` function is convenient. We can describe all of the variables in a data set (a "dataframe" in R) using the function on our dataframe object, dat:
```{r}
summary(dat)
```
We can apply to a single variable:
```{r}
summary(dat$TchrRating)
```
We can also perform functions like `min()`, `max()`, `mean()` and assign them to objects that we can later use:
```{r}
minX <- min(dat$TchrRating)
maxX <- max(dat$TchrRating)
meanX<- mean(dat$TchrRating)
```
The script above just assigned the output of the functions to objects. Let's view the objects in the console:
```{r}
meanX
minX
maxX
```
We can get the variance of a variable using the 'var()' function:
```{r}
var(dat$TchrRating)
```











