# Simple Analyses of Complex Survey Data {#Complex}

In a garden variety simple-random-sample design, we randomly sample units from a larger population and then estimate the mean, standard deviation, and confidence intervals to subsequently draw an inference about the larger population. This is the assumption underlying the example and explanations in the previous chapters. In this chapter we examine complex survey data. In survey designs with complex data, we need to make corrections to the standard error, which in turn affects the confidence intervals, which subsequently affects our inference about the statistic---such as the mean---in the population we seek to make claims about.

## Packages
There are two packages that are handy for analyzing complex survey data: The survey package [@R-survey] and the similarly named srvyr package [@R-srvyr], which works with the survey package to make the output easier to work with (because why would the survey package on its own make life easy?). Harkening back to a previous chapter, we need to first be sure we have these packages installed. We will also use the dplyr package [@-dplyr] again, as this uses the same data-wrangling language that the srvyr package does.^[Whenever we use a package, it is a good idea to cite it. People work hard to create these (well usually---for some, it is disgustingly easy). To get the citation, we can use the `citation()` function, usually, as in `citation("srvyr")`. ]

```{r warning = FALSE, message = FALSE}
library(survey)
library(srvyr)
library(dplyr)
```

## Example data
To go through this example, let's use the data set that is presented in the chapter by Heeringa, West, and Berglund [-@heeringa_applied_2010, Chapter 2], who provide an explanation that we can follow along with as we actually perform these calculations.^[In their chapter, their variable of interest is persons' ratings of the economy. We use the same data but for demonstration call the outcome variable _scores_ and label the cases _students_, who are nested within the school clusters, which are in districts, serving as strata. For this demo, we will ignore weighting.]


```{r}
dat <- read.csv("Heeringa.Complex.SampleData.csv")
str(dat)
head(dat)
```

## The three hypothetical complex designs

### Pretending we have a cluster-sampled design
We will use this data set to practice analyzing the data in a cluster-sampled design. This would be the kind of analysis we would conduct if we had randomly sampled schools---that is, our clusters---from a broader population of schools (such as from a sampling frame that listed all of the schools in the country). In this design, our _primary sampling unit_ is schools. Within each school, we have students, whom we had administered the survey to.^[Before we got to this step, we had previously calculated each student's score, which is the `score` variable in our data set. This score might be a composite score calculated as the mean (or sum) across several items, as was demonstrated in the previous chapter.] This design is very convenient because it is easier to randomly sample schools from a list of schools, whereas it is virtually impossible to have a list of all of the individual students in the country. We pay a price for this convenience, though, because students who are nested within schools tend to live in similar neighborhoods, share the same values, have the same learning experiences, and so forth, so if we used a run-of-the-mill standard error (and confidence interval), we will be overly optimistic in how precise we are in estimating the population. When we correctly analyze these data, we pay a penalty because our standard error is corrected to be larger than it would be with a simple random sample.

### Pretending we have a stratified design
After analyzing the cluster-sampling design data, we'll erase our memories of that design and pretend we're analyzing a stratified design. In this type of design, we have have conducted a simple random sample of students within strata. In this case, the students are our _primary sampling unit_. Let's say we have a list of school districts---these are our strata. We then have a list of all the students in each district, which is our sampling frame. We randomly selected students from each district, then we estimate the mean of the population. This is a very powerful design if we believe that the strata explain a lot of variation in the variable (in the score, in this case). It helps us to have greater precision in our estimate of the population mean.^[Strata serve as a kind of fixed effect or a blocking variable in ANOVA.]


### Pretending we have a stratified cluster-sampled design
After monkeying around with a stratified design, we wipe our memories of how we had designed that study and pretend we had collected the data using schools as our primary sampling unit, similar to how we did with the cluster-sampled design, but this time we sample schools from within their districts. The districts are strata. So, we have students nested within schools, which were randomly selected, which in turn are nested within strata. This is a very practical and smart design to use. First, we have the convenience of sampling schools, which is far easier than randomly sampling students from lists. Second, we can reduce that penalty to the standard error of this clustering by also using a stratification variable---the districts. This allows us to make claims about the population of schools within these particular districts. 

We will also pretend we have a simple random sample, as well, to compare these scenarios with each other in terms of the standard error and confidence intervals.

## Cluster-sampled design

### Preparing the data
In the example that Heeringa and colleagues provide, they did not use a finite population correction. Another way to think about this is that the population of our primary sampling unit is very large. Let's say, therefore, that there is a very large number of schools (such as one million) in the population.

For many researchers, they treat the population as an infinitely large number, so specifying the population as very large is a conservative approach. For other researchers, such as for evaluations in a specific population---such as the schools in a particular state---the population size is finite.

The following example treats the population as infinite (to follow Heeringa's example), but in your own study, you can specify these sizes to match your needs in your research or evaluation. 

To get this to work with the survey package, we create a column in the data that is this cluster _N_-size. Here, we can use `N_cl` as this column name to remind us that this is the population size of clusters.

```{r}
dat$N_cl <- 1000000
head(dat)
```


### Step 1: Specifying the design
To use the srvyr package (and the survey package it wraps around), we need to create a survey-design object.^[For more information on this package, we can use `help(package = "srvyr")` or https://cran.r-project.org/web/packages/srvyr/vignettes/srvyr-vs-survey.html.] This is our way of saying with the survey design was when we collected the data. 

Let's create a survey-design object, with which we can perform subsequent calculations. We do this with the srvy package's `as_survey_design()` function. To remind us that this is a design and that it uses clusters, we can name our object `des_cl`, but you could call it `Bill` or `mydesign`.

```{r warning=FALSE, message = FALSE}
library(survey)
library(srvyr)
des_cl <- dat %>% 
           as_survey_design(id = school, 
                           fpc = N_cl)
```

The `id =` specifies what the primary sampling unit is. The primary sampling unit (PSU) is the school variable in our data set. We would think that the `fpc =` argument specifies the finite population correction, but it does NOT! Unintuitively, the default in the survey package is for `fpc` to actually be the *size* of the population of primary sampling units. So, this should be called something like `pop` instead of `fpc` in the program---that would make it too easy, though, wouldn't it.^[But we can also use the `fpctype = "correction"` argument to specify the type of fpc discussed in the Heeringa chapter (which was 1 - _n_/_N_). Unfortunately, although we get the correct design effect, the estimated standard error is wrong.] If the number of clusters in our population were finite, like 168 schools, we can use the `fpc = 168` as a value for each case in the data set, which would make our standard error smaller. However, for this hypothetical example, we are saying there are 1 million schools.

We can look at the survey design we created using the `summary()` function on that object:

```{r}
summary(des_cl) 
```
Notice that the probability of being sampled is equal across the quantiles. This is because everyone has the same fpc. 

### Step 2: Calculating the estimates
We want to know the mean and variance of our sample, as well as the confidence interval of the mean. We cannot use the regular Base R functions like `mean()` and `variance()` because they do not correct for the design. Instead, we use the functions from the svyr package. For example, instead of `mean()`, it uses `survey_mean()`.^[We're using piping, `%>%` with this particular package just like how we did with the dplyr package. These are part of the tidyverse community of R packages.]

```{r}
des_cl %>% 
  summarize(survey_mean(score,                                 
                        vartype = c("se","ci"), df = degf(.),  
                        deff = TRUE ))
```

We have the `summarize()` function around the `survey_mean()` function. Just as we do in mathematics, we work from inside the parentheses to the outside. The first argument in the `survey_mean()` function is the variable that we are estimating, `score` in this case. The second argument specifies the type of sampling-distribution variance we would like to report (not to be confused with the sampling variance). Here, we seek the standard error (`"se"`) and the upper and lower 95% confidence interval (`"ci"`) based on the _t_ distribution (based on the `df = degf(.)` argument. If we wanted the _z_ distribution, we would remove the `df = ` argument. We also request the design effect to be reported, `deff = TRUE`.

The coefficient of interest is the mean. The standard error and the lower- and upper-boundaries of the confidence intervals are reported, as well as the design effect.

We can actually save this output to an object and then use it later on. I don't like how outputs get printed from the srvyr. I want to see two decimal places, so I'll round it to two decimal places, then print it.^[We can also use RStudio's `View()` function to see it in an extra tab, with `View(rslts_M_cl)`.]

```{r}
rslts_M_cl <- des_cl %>% 
          summarize(Mn = survey_mean(score, 
                         vartype = c("se","ci"), df = degf(.), 
                         deff = TRUE ) )
print.data.frame( round(rslts_M_cl, 2) )
```
Now let's examine the variance and the sampling distribution of the variance.
The function is `survey_var()`:

```{r}
rslts_V_cl <- des_cl %>% summarize(survey_var(score) )
print.data.frame( round(rslts_V_cl, 2) )
```
We can perform both the `survey_mean()` and `survey_var()` and save these to a report:

```{r}
rslts_cl <- des_cl %>% 
          summarize(Mn = survey_mean(score, 
                         vartype = c("se","ci"), df = degf(.), 
                         deff = TRUE ),
                    Var= survey_var(score) )
print.data.frame(round(rslts_cl, 2))
```
So, we have an estimated mean on the score---whatever construct this happens to be---of $`r round(rslts_cl$Mn ,2)`$ and a 95% confidence interval of $`(r round(rslts_cl$Mn_low ,2)`, `r round(rslts_cl$Mn_upp ,2)`)$. 


### Effective sample size
We can calculate the effective sample size from the number of students within the clusters, _n_, and the design effect. This is helpful for explaining the design effect to folks such as school personnel. First, let's get the _n_-size of students. 

To specify what the _sample_ size of our smallest units (students) is, we can count the number of rows in the data using `nrow(dat)`. This is useful if each row is an individual person. We can also use `length(unique(dat$sid))` to count the number of observations that are unique in the student ID column.^[Sometimes data structures, such as those used in longitudinal analyses, have more than one row for each person. The `unique()` function makes sure we don't count somebody with the same ID twice.] Either way, we get the same number.

```{r}
n <- length(unique(dat$sid)) 
n
```

As expected, there are 32 units (students) in our sample. Here's the effective sample size, which is calculable as $\frac{n}{\text{design effect}}$.

```{r}
rslts_cl <- rslts_cl %>% mutate(EffSS = n/Mn_deff)
print.data.frame(round(rslts_cl, 2))
```

The standard error with this design is about what it would be with a simple random sample of $`r round(rslts_cl$EffSS ,2)`$ (or rounding up, $`r ceiling(rslts_cl$EffSS)`$) students from the population. This shows the penalizing effect that we have correctly applied to the data.

### Reporting the mean per cluster
Sometimes we want to also report the mean per school. We can use
the `group_by()` function to specify what the grouping is before we apply the functions. Here, we're grouping by school, then estimating the mean and variance^[This function is part of the tidyverse language in R, which is what the dplyr and srvyr packages use.]:

```{r}
M_by_schl <- des_cl %>% 
              group_by(school) %>% 
                summarize(Mn = survey_mean(score),
                          Var= survey_var(score))
print.data.frame(round(M_by_schl,2))
```

### Two-stage cluster-sampled design
The method so far has assumed that the units within the clusters were the population within those clusters. For example, if we surveyed all of the students in our selected schools. In a two-stage design, we have the extra step of randomly selecting units within the clusters.

We first have to prepare the data. Let's pretend we had a _population_ of 100 students within each school from which we had randomly sampled. It is super unlikely that we'd ever have a data set with the same population size in all of the schools. This is only a demonstration---in our actual schools, we will probably know the total number in each school. There are big schools and small schools. 

```{r}
dat$N_cl_s <- 100
```

Now, we can use this column in the data set.

```{r}
des_cl2 <- dat %>% 
               as_survey_design(id = c(school, sid), 
                               fpc = c(N_cl, N_cl_s))
summary(des_cl2) 
```

```{r}
rslts_cl2 <- des_cl2 %>% 
          summarize(Mn = survey_mean(score, 
                         vartype = c("se","ci"), df = degf(.), 
                         deff = TRUE ),
                    Var= survey_var(score) )
rslts_cl2 <- rslts_cl2 %>% mutate(EffSS = n/Mn_deff)
print.data.frame(round(rslts_cl2,2))
```
The results are not that different from the single-stage cluster-sample design we just analyzed. Again, this is an artificial data set and we would likely specify `N_cl_s` for each person within a school the same but different across schools (of different sizes).


## Stratified design
The strata are districts. In this sampling design, we are randomly selecting students from each stratum and pretending that the school cluster did not exist when we designed and carried out this sampling plan.

```{r}
head(dat)
```
Again, we'll ignore the finite population correction and assume that the population is big. This yields a more conservative confidence interval than if we used a finite population. Here, our primary sampling unit (PSU) is the lowest element of our sampling plan. In other words, our PSU is the students. The Survey package still includes an argument for the PSU (`id = `), and instead of clusters, we specify `id = ~1`, where 1 tells R to treat the individual cases as the PSU.

Traditionally, _h_ stands for stratum. So, we'll call our design object `des_h`. 

We need to specify how many sampled units are in each stratum. In other words, how many students are in each district. Normally, there will be a different number of students in a district (some districts are large; some are small). Here, for convenience (and to replicate the results in Heeringa and colleagues' chapter), we will make them all very large, maybe 25,000,000 students in each district. In the column in the data, we can think of the students as coming from their district of this population size.

```{r}
dat$N_h   = 25000000
```

### Step 1: Specifying the design

```{r}
des_h <- dat %>% 
  as_survey_design(id = 1, 
                   strata = district,
                   fpc = N_h)
  
summary(des_h) 
```
Notice that the population size is number of students per stratum. Again, in our example, we're saying each stratum has the same number of students, but if we had a different population size in each stratum, we could specify that `data$N_h` to be that number within each stratum (e.g., Manuel and all of the other students in his district come from a small district but Julio and his colleagues come from a big one).

### Step 2: Calculating the estimates

This step is identical to what we did before with cluster-sample design. The difference is the design itself, which we created in the previous step.
```{r}
rslts_h<- des_h %>% 
        summarize(Mn = survey_mean(score, 
                       vartype = c("se","ci"), df = degf(.), 
                       deff = TRUE ),
                  Var= survey_var(score) )
rslts_h
```

Here's the effective sample size from _n_ and the design effect.
First, let's remember how many observations are in our data:

```{r}
n
rslts_h <- rslts_h %>% mutate(EffSS = n/Mn_deff)
rslts_h
```
As an extra step, we could report the mean per stratum:
```{r}
M_by_dist <- des_h %>% 
             group_by(district) %>% 
               summarize(Mn = survey_mean(score),
                         Var= survey_var(score))
M_by_dist
```


## Stratified design

### Step 1: Specifying the design
Now, we specify the PSU as the cluster (which are schools)
and the strata as the districts. And again, we use a very low
finite population correction (fpc) again to estimate the std error
as if the population were huge (i.e., n/N = 0.00001).

```{r}
#********************************************
# Complex Sampling Design: A Stratified Cluster Randomized Design
#------------
# Step 1: Create a survey-design object.



dat$N_hcl <- 1000000


des_hcl <- dat %>% 
  as_survey_design(id = school,
                   strata = district,
                   fpc = N_hcl)

summary(des_hcl) 
# Notice that the population size is the number of clusters per
# stratum. In other words, we're saying there are 250,000 schools
# in each stratum. That's a lot. However, this is the spirit of 
# model-based inference, where the population is more theoretical.

#------------
# Step 2: Calculate estimates using functions on the design.
# Mean:

rslts_hcl <- des_hcl %>% 
        summarize(Mn = survey_mean(score, 
                       vartype = c("se","ci"), df = degf(.), 
                       deff = TRUE ),
                  Var= survey_var(score) )

# Effective Sample Size from n and the design effect. 
# First, let's remember how many observations are in our data:
n

rslts_hcl <- rslts_hcl %>% mutate(EffSS = n/Mn_deff)
rslts_hcl

#------------
# Extra step: 
# We can look at the mean per stratum :

M_by_Dist_cl <- des_hcl %>% 
             group_by(district) %>% 
               summarize(Mn = survey_mean(score),
                         Var= survey_var(score))
M_by_Dist_cl


#********************************************
# Simple Random Sample Design: 
# Assuming that students were randomly drawn from the population
#------------

des_srs <- dat %>% 
               as_survey_design(id = 1, fpc = N)

rslts_srs <- des_srs %>% 
        summarize(Mn = survey_mean(score, vartype = c("se","ci"), df = degf(.)),
                  Var= survey_var(score)) 

rslts_srs <- rslts_srs %>% 
               mutate(EffSS   = n,
                      Mn_deff = 1)
rslts_srs

#********************************************************
# Comparing all design scenarios
#********************************************************
# Note that we would NOT do this in practice.
# In practice, we would select a design and report 
# the results.
rslts_cl2
rslts_cl
Designs_report       <- bind_rows(rslts_srs, rslts_cl,rslts_cl2, rslts_h, rslts_hcl)
Designs_report$Design <- c("Simple Rand. Samp.", "Clustered", "two-stage Clustered", "Stratified", "Stratified Clustered")

Designs_report %>% select(Design, Mn, Mn_se, Mn_low, Mn_upp, Mn_deff, EffSS, Var, Var_se)


# Compare this to Table 2.2 in Heeringa, et al. The small differences
# (e.g., in the effective n-size) are due to rounding error in the chapter.
           

# Write it to csv file if you want to view it in Excel:                         
# write.csv(Designs_report, "Designs_report.csv", row.names = FALSE)




                         
```







### Step 2: Calculating the estimates








***
Updated on `r Sys.Date()`.

***










